---
title: "Assignment 1"
author: "Text as Data"
date: "2023-09-18"
output: 
  pdf_document:
    keep_tex: true
    citation_package: natbib
header-includes:
- \usepackage{booktabs}
- \usepackage{xcolor}

extra_dependencies: ["hyperref","booktabs"]
urlcolor: blue
bibliography: ../presentation-resources/MyLibrary.bib
---

# Introduction

This first assignment will test your grasp of some of the concepts and methods we covered in sessions 1-2. Give an answer to each question, even if you are not sure. In an ideal world, you should present your work as a document with integrated code, execution and text, such as an Rmd file, or an .ipynb notebook, but if this doesn't work out a script and a pdf will do fine.

# 1 

Consider the two following texts.

```{r echo=TRUE, include=TRUE}
texts <- c(
  "I don't like cricket",
  "You like cricket"
)
```

1.1 Write two text processing pipelines that create document feature matrices that **do** and **do not** preserve the differences in the texts. 

1.2. Describe how the preprocessing choices you make affect the representation of those texts.

1.3. Give one example each of **tasks** where your text preprocessing choices that do not preserve the differences in the texts **would** and **would not** limit our ability to perform the task.

# 2

Now consider the following three texts.

```{r echo=TRUE, include=TRUE}
texts <- c(
  "Climatic change is causing adverse impacts",
  "Changes in the climate have caused impacts to human systems",
  "Chelsea have a goal difference of zero in the premier league this season"
)
```

2.1. Turn these texts into a document feature matrix without any additional pre-processing steps

2.2. Calculate a simplistic measure of how similar each text is to each other, by reporting the number of columns where both texts contain a non-zero value (you can do this with code or by hand).

2.3. Create a text processing pipeline that results in a matrix that preserves the similarity  we can see intuitively between the texts.

2.4. Comment on why the additional pre-processing steps created a more useful representation of the texts *in this case*.

